Parameter Inference\\with Bifurcation Diagrams

### 2021-05-23 docs/article.tex

\begin{figure}
    \centering
    \includegraphics[width=7cm]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(s)$ and determinant $\Det$ along two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars}
    \label{fig:measure}
\end{figure}

In order for predicted bifurcations $p(\theta)\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. We expect this term to be some mean over the targets $\targets$ and predictions $\predictions(\theta)$. We choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target in cases where $|\predictions|\geq|\targets|$. This term only zero when each target is matched by at least one prediction
\begin{align}
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:error}
\end{align}
We can see from Figures \ref{fig:minimal-models} and definitions \eqref{eq:predictions} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$. Using these quantities we can define a positive semi-definite measure $\measure(u,p)$ of zero crossings in the determinant along the bifurcation curve. Figure \ref{fig:measure} demonstrates this measure is maximal at bifurcation at locations $p(\theta)$ and has finite gradients in non-bifurcating regimes. The total bifurcation measure $\Psi(\theta)$ is defined as
\begin{align}
    \Psi(\theta):=\frac{
        \int_{\rates(u,p)=0}\!
        \measure(u,p)\,\mathrm{d}u\mathrm{d}p
    }{
        \int_{\rates(u,p)=0}\!\
        \mathrm{d}u\mathrm{d}p
    }
    \quad\mathrm{where}\quad
    \measure(u,p):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{align}
Here we denote $\int_{\rates(u,p)=0}\mathrm{d}u\mathrm{d}p$ as the sum of the line integrals in $(u,p)\in\Reals^{N+1}$ defined by the level set $\rates(u,p)=0$. The measure $\measure(u,p)$ is one at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in the directional derivative $\frac{d}{ds}\Det$ going to zero and hence the measure $\measure(u,p)$ going to zero for each turning point. While the calculation of the determinant is straightforward, its directional derivative is not. Fortunately, a tangent field representation of the bifurcation curve allows us to compute this derivative implicitly; see Appendix \ref{appendix:tangent-fields} for details.

As the determinant $\Det$ diverges we approach regimes far away from any closest bifurcation and hence $\measure(u,p)\rightarrow0$. We would still like to have non-zero gradients with respect to $\theta$ in this regime, and therefore $\measure(u,p)$ was designed to go to zero slowly. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the controlled condition region $p$ is densely packed with bifurcations. The measure $\measure(u,p)$ is also invariant under certain classes of transformations on $\rates(u,p)$ which do not change bifurcation locations, proofs of which are provided in Appendix \ref{appendix:measure}. The total measure $\Psi(\theta)$ is added to the supervised term as if it were a likelihood; this defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\,\,\lambda \log\Psi(\theta)+
    \frac{1}{|\targets|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:loss}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached. Note that in while individual bifurcations $p(\theta)$ depend smoothly on $\theta$, the total number of predictions $|\predictions|$ does not have gradient contributions with respect to $\theta$. We can safely drop the dependency in the prediction counter and now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ and total measure $\Psi(\theta)$
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient-loss}
\end{align}
In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation curve $\rates(u,p)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:space-curve}.

@@

\subsection{Minimal Models}
Figures \ref{fig:saddle-node:results} and \ref{fig:pitchfork:results} show example optimisations of $(\theta_1,\theta_2)$ for the minimal saddle-node and pitchfork models respectively. The magnitude of the correction factor $\lambda=0$ when bifurcations are present and $\lambda\neq0$ otherwise. Optimisation trajectories approach lines of global minima in the cost function, which represent a set of geometrically equivalent models, depicted in the right panels. Two bifurcation curves are geometrically equivalent if the number, type and locations of bifurcations match.

We can see that the geometrically equivalent lines are contained within larger basins defined by $\lambda=0$ where the correct number and type of bifurcations are present $p\in\Omega$, but do not match the locations of targets $\targets$. All models within this basin are in some sense topologically equivalent. This hierarchical classification allows us to identify the set of models that satisfy observed qualitative behaviour \cite{Stumpf2019ParameterBifurcations} before any attempt at inferring kinetic parameters, which is done by choosing a model along the line of geometrically equivalent models.

@@

\begin{figure}
\centering
\includegraphics[width=6cm]{saddle-landscape.png}
\includegraphics[width=6cm]{saddle-optima.png}
\caption{Saddle-node model $\rates(u,p) = p + \theta_{1}u+\theta_{2}u^3$ optimised with respect to targets $\targets=\{-1,1\}$ in observation region $\Omega\in[-2,2]$. The right panel shows bifurcations diagrams for the three optima marked by stars on the left panel. The optimisation trajectories in white follow the gradient of the cost, approaching the black line of global minima in the left panel}
\label{fig:saddle-node:results}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=6cm]{pitchfork-landscape.png}
\includegraphics[width=6cm]{pitchfork-optima.png}
\caption{Pitchfork model $\rates(u,p) = \theta_{1} + u p +\theta_{2}u^3$ optimised with respect to targets $\targets=\{0\}$ in observation region $\Omega\in[-5,5]$. The right panel shows bifurcations diagrams for the three optima marked by stars on the left panel. The optimisation trajectories in white follow the gradient of the cost, approaching the black lines of global minima in the left panel}
\label{fig:pitchfork:results}
\end{figure}

\subsection{Genetic Toggle Switch}
In this section we optimise a model where the states share a Hill function relationship with cooperatively $n=2$; these models often emerge from mass action kinetics and are used to model species concentrations. After re-scaling the equations governing the dynamics of concentrations, the simplified equations for state $u_$ and $u_2$ become 
\begin{equation}
    \partial_t u_1 = \frac{ a_1 + (p u_2)^2}{ 1 + (p u_2)^2 } - \mu_1 u_1 \quad
    \partial_t u_2 = \frac{ a_2 + (k u_1)^2}{ 1 + (k u_1)^2 } - \mu_2 u_2
    \label{eq:two-state}
\end{equation}
where $a_k$ are 
 the  we can recover both activator and inhibitor relationships.
Figure \ref{fig:two-state-optima} shows a UMAP embedding of the optimal parameter basins, represented as point clouds. 16 distinct clusters are obtained using DBSCAN. Each cluster corresponds to a kinetically different model regions that are qualitatively equivalent. See Appendix \ref{appendix:clusters} for models that represent the centroid of each cluster.

>>>>

\begin{figure}
    \centering
    \includegraphics[width=7cm]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(s)$ and determinant $\Det$ along the arclength $s$ of two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars}
    \label{fig:measure}
\end{figure}

In order for predicted bifurcations $p(\theta)\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. We expect this term to be some mean over the targets $\targets$ and predictions $\predictions(\theta)$. We choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target in cases where $|\predictions|\geq|\targets|$. This term only zero when each target is matched by at least one prediction
\begin{align}
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:error}
\end{align}
We can see from Figures \ref{fig:minimal-models} and definitions \eqref{eq:predictions} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$. Using these quantities we can define a positive semi-definite measure $\measure(s)$ of zero crossings in the determinant along the bifurcation curve. Figure \ref{fig:measure} demonstrates this measure is maximal at bifurcation at locations and has finite gradients in non-bifurcating regimes. The total bifurcation measure $\Psi(\theta)$ over a diagram is defined as
\begin{align}
    \Psi(\theta):=\frac{
        \int_{\rates(u,p)=0}\!
        \measure(s)\,\mathrm{d}s
    }{
        \int_{\rates(u,p)=0}\!\
        \mathrm{d}s
    }
    \quad\mathrm{where}\quad
    \measure(s):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{align}
Here we denote $\int_{\rates(u,p)=0}\mathrm{d}s$ as the sum of the line integrals in $(u,p)\in\Reals^{N+1}$ defined by the level set $\rates(u,p)=0$ with $s$ being an arbitrary parametrisation of the curves. The measure $\measure(s)$ is one at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in the directional derivative $\frac{d}{ds}\Det$ going to zero and hence the measure $\measure(s)$ going to zero for each turning point. While the calculation of the determinant is straightforward, its directional derivative is not. Fortunately, a tangent field representation of the bifurcation curve allows us to compute this derivative implicitly; see Appendix \ref{appendix:tangent-fields} for details.

As the determinant $\Det$ diverges we approach regimes far away from any closest bifurcation and hence $\measure(s)\rightarrow0$. We would still like to have non-zero gradients with respect to $\theta$ in this regime, and therefore $\measure(s)$ was designed to go to zero slowly. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the controlled condition region $p$ is densely packed with bifurcations. The measure $\measure(s)$ is also invariant under certain classes of transformations on $\rates(u,p)$ which do not change bifurcation locations, proofs of which are provided in Appendix \ref{appendix:measure}. The total measure $\Psi(\theta)$ is added to the supervised term as if it were a likelihood; this defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\,\,\lambda \log\Psi(\theta)+
    \frac{1}{|\targets|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:loss}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached. Note that in while individual bifurcations $p(\theta)$ depend smoothly on $\theta$, the total number of predictions $|\predictions|$ does not have gradient contributions with respect to $\theta$. We can safely drop the dependency in the prediction counter and now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ and total measure $\Psi(\theta)$
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient-loss}
\end{align}
In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation diagram $\rates(u,p)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:leibniz-rule}.

@@

\subsection{Minimal Models}
Figures \ref{fig:minimal-models:results} show example optimisations of $(\theta_1,\theta_2)$ for the minimal saddle-node and pitchfork models respectively. Optimisation trajectories approach lines of global minima in the cost function, which represent a set of geometrically equivalent models, depicted in the right panels. Two bifurcation diagrams are geometrically equivalent if the number, type and locations of bifurcations match the specified targets $\targets$.

We can see that the geometrically equivalent lines are contained within larger basins where the correct number and type of bifurcations are present but do not match the locations of targets $\targets$. All models within this basin are in some sense topologically equivalent. This hierarchical classification allows us to identify the set of models that satisfy observed qualitative behaviour \cite{Stumpf2019ParameterBifurcations} before any attempt at inferring kinetic parameters, which is done by choosing a model along the line of geometrically equivalent models.

@@

\begin{figure}
\centering
\includegraphics[width=6cm]{saddle-landscape.png}
\includegraphics[width=6cm]{saddle-optima.png}
\includegraphics[width=6cm]{pitchfork-landscape.png}
\includegraphics[width=6cm]{pitchfork-optima.png}
\caption{Saddle-node $\rates(u,p) = p + \theta_{1}u+\theta_{2}u^3$ and pitchfork $\rates(u,p) = \theta_{1} + u p +\theta_{2}u^3$ optimised with respect to $\theta$ so that predicted bifurcations $\predictions(\theta)$ match targets $\targets$ in control condition $p$. The right panel shows bifurcations diagrams for the three optimal $\theta^*$ marked by stars on the left panel. The optimisation trajectories in white follow the gradient of the cost, approaching the black lines of global minima in the left panel}
\label{fig:minimal-models:results}
\end{figure}

\subsection{Genetic Toggle Switch}
In this section we optimise a model where the states share a Hill function relationship with cooperatively $n=2$; these models often emerge from mass action kinetics with quasi-steady state approximations and are used to model species concentrations. After re-scaling the equations governing the dynamics of concentrations, the simplified equations for state $u_1$ and $u_2$ become 
\begin{equation}
    \partial_t u_1 = \frac{ a_1 + (p u_2)^2}{ 1 + (p u_2)^2 } - \mu_1 u_1 \quad
    \partial_t u_2 = \frac{ a_2 + (k u_1)^2}{ 1 + (k u_1)^2 } - \mu_2 u_2
    \label{eq:two-state}
\end{equation}
where $a_k$ is the baseline production rate for species $k$ in the absence of the other species. Each species has a finite degradation rate $\mu_k$. Finally we have two half-saturation constants $p$ and $k$, one of which is chosen as our control condition. A baseline production rate $a_k>1$ recovers an inhibitor type hill function for species $k$ and is an activator otherwise. The half-saturation are proportional to the slope of the hill activations.

%  the  we can recover both activator and inhibitor relationships.
% Figure \ref{fig:two-state-optima} shows a UMAP embedding of the optimal parameter basins, represented as point clouds. 16 distinct clusters are obtained using DBSCAN. Each cluster corresponds to a kinetically different model regions that are qualitatively equivalent.
